# Time Series Data

**Raw data**: The raw values without any transformation. We are not so
interested in the raw data, as it is hard to read information from it.
Take the GDP plot as an example (@fig-gdpts, upper-left subplot). There
is an overall upward trend. But we are more interested in: how much does
the economy grow this year? Is it better or worse than last year? The
answers are not obvious from the raw data. Besides, there are obvious
seasonal fluctuations. Usually the first quarter has the lowest value in
a whole year, due to the Spring Festival, which significantly reduces
the working days in the first quarter. The seasonal fluctuations
prohibit us from sensibly comparing two consecutive values.

**Growth rate**: The headline GDP growth is usually derived by comparing
the current quarter with the same quarter from last year.
$g=\frac{x_t - x_{t-4}}{x_{t-4}}\times 100.$ This makes sense. As
mentioned above, due to seasonal patterns, comparing two consecutive
quarters directly does not make sense. The year-on-year growth rate
directly tells us how fast the economy grows. However, by dividing the
past values, it loses the absolute level information. For instance, it
is hard to tell after the pandemic, whether or not the economy recovers
from its pre-pandemic output level. Besides, it is sensitive to the
values of last year. For example, due to the pandemic, the GDP for 2020
is exceptionally low, which makes growth rate for 2021 exceptionally
high. This is undesirable, because it does not mean the economy in 2021
is actually good. We would like a growth rate that shirks off past
burdens.

That's why we sometimes prefer (annualized) quarterly growth rate.
$g=\frac{x_t-x_{t-1}}{x_{t-1}}\times 400.$ Due to seasonally patterns,
two consecutive quarters are not comparable directly. A first quarter
value is usually much lower than the fourth quarter of last year due to
holidays, which does not necessarily mean the economy condition is
getting worse. Since this pattern is the same every year, it is possible
to remove the seasonal fluctuations. This is called *seasonally
adjustment*. We won't cover seasonally adjustment in detail, but the
next section will give some intuitions on how this can possibly be done.
After seasonally adjusting the time series, we can calculate the growth
rate based on two consecutive values (annualized by multiplying $4$).
The bottom-right panel of @fig-gdpts is the seasonally-adjusted
quarterly growth. Note that it is no longer biased upward in 2021 as the
YoY growth.

**Seasonally-adjusted series**: This is usually the data format we
prefer in time series analysis. FRED reports both seasonally-adjusted
and non-seasonally-adjusted series. Seasonal adjustment algorithm is a
science in itself. Popular algorithms include X-13-ARIMA developed by
the United States Census Bureau, TRAMO/SEATS developed by the Bank of
Spain, and so on.

```{r}
#| include: false
library(zoo)
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Quarterly GDP Time Series (Unit: RMB Billion or %)"
#| label: fig-gdpts
data = readRDS("data/gdp.Rds")
par(mfrow=c(2,2), mar=rep(2,4))
ts.plot(data$GDP)
ts.plot(data[, c("GDP","GDPSA")], col=1:2)
ts.plot(data$YoY)
ts.plot(window(data$QoQ, start=2011, end=2022))
```

**Log levels and log growth rates**: We like to work with log levels. A
lot of economic time series exhibit exponential growth, such as GDP.
Taking logs convert them to linear. Another amazing thing about logs is
the difference of two log values can be interpreted as percentage
growth. We know from Taylor expansion that for small values of
$\Delta x$ : $\ln(\Delta x +1) \approx \Delta x$. Therefore,

$$
\ln x_t - \ln x_{t-1} = \ln\left(\frac{x_t}{x_{t-1}}\right) = \ln\left(\frac{x_t-x_{t-1}}{x_{t-1}}+1\right) \approx \frac{x_t-x_{t-1}}{x_{t-1}}.
$$

So it is very handy to just difference the log levels to get the growth
rates. Log difference can also be interpreted as the continuously
compounded rate of change, if assuming

$$
\frac{x_t}{x_{t-1}}=e^g \implies g = \ln x_t - \ln x_{t-1}.
$$

Log difference also has the property of summability: summing up a series
of log differences gives the log level provided the initial level. It is
not as handy if you want to recover the level values from a series of
percentage growth.

$$
\ln x_t = x_0 + \sum_{j=1}^{t} (\ln x_j - \ln x_{j-1}).
$$

::: callout-tip
## Exercise

Buying vs. renting a home, which is better? Compute the NPV:

$$
\text{NPV}=\sum_{t=0}^T \frac{C_t}{(1+r)^t}=\int_0^T C(t) e^{-rt} dt.
$$
:::
