# Time Series Data

### Raw data {.unnumbered}

We are not so interested in the raw data without any transformation, as
it is hard to read information from it. Take the GDP as an example
(@fig-gdpts, upper-left). There is an overall upward trend. But we are
more interested in: how much does the economy grow this year? Is it
better or worse than last year? The answers are not obvious from the raw
data. Besides, there are obvious seasonal fluctuations. Usually the
first quarter has the lowest value in a whole year, due to holidays that
significantly reduce the working days in the first quarter. But this
does not necessarily mean the economic condition of the first quarter is
worse than others. The seasonality prohibits us from sensibly comparing
the values of two consecutive quarters.

### Growth rates {.unnumbered}

The headline GDP growth rates are usually reported by comparing the
current quarter with the same quarter from last year:
$g=\frac{x_t - x_{t-4}}{x_{t-4}}\times 100.$ As mentioned above, due to
seasonal patterns, comparing two consecutive quarters does not make
sense. The year-on-year growth rate tells us how fast the economy grows.
However, it loses the information about absolute levels. That comes with
sereval drawbacks. For instance, it is hard to tell whether the economy
recovers to the pre-pandemic level after the shock. Due to the
unprecedented impact of the pandemic, the GDP for 2020 is exceptionally
low, which renders the growth rate for 2021 artifically high. This is
undesirable, because it does not mean the economy in 2021 is actually
good. We would like a growth rate that shirks off the excessive
influence of past observations.

That's why we sometimes prefer (annualized) quarterly growth rate:
$g=\frac{x_t-x_{t-1}}{x_{t-1}}\times 400.$ Due to seasonally patterns,
two consecutive quarters are not comparable directly. But, since this
pattern is the same every year, it is possible to remove the seasonal
fluctuations. This is called *seasonally adjustment*. We will not cover
the seasonal adjustment method here. But this is something that can be
done. After seasonally adjusting the data, we can calculate the growth
rate based on two consecutive values (annualized by multiplying $4$).
The bottom-right panel of @fig-gdpts shows the seasonally-adjusted
quarterly growth. Note that it is no longer biased upward in 2021 as the
year-on-year growth rate.

### Seasonally-adjusted data {.unnumbered}

This is usually the data format we would prefer in time series analysis.
[FRED](https://fred.stlouisfed.org/) reports both seasonally-adjusted
and non-seasonally-adjusted series. The method for seasonal adjustment
is a science in itself. Popular algorithms include X-13-ARIMA developed
by the United States Census Bureau, TRAMO/SEATS developed by the Bank of
Spain, and so on.

```{r}
#| include: false
library(zoo)
```

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Quarterly GDP Time Series (Unit: RMB Billion or %)"
#| label: fig-gdpts
data = readRDS("data/gdp.Rds")
par(mfrow=c(2,2), mar=rep(2,4))
ts.plot(data$GDP)
ts.plot(data[, c("GDP","GDPSA")], col=1:2)
ts.plot(data$YoY)
ts.plot(window(data$QoQ, start=2011, end=2022))
```

### Logarithms {.unnumbered}

We like to work with log values. Lots of economic time series exhibit
exponential growth, such as GDP. Taking logs convert them to linear.
Another amazing fact about logs is that the difference between two log
values can be interpreted as percentage growth. To see this, by Taylor's
theorem, we have $\ln(\Delta x +1) \approx \Delta x$ for small values of
$\Delta x$. Therefore,

$$
\ln x_t - \ln x_{t-1} = \ln\left(\frac{x_t}{x_{t-1}}\right) = \ln\left(\frac{x_t-x_{t-1}}{x_{t-1}}+1\right) \approx \frac{x_t-x_{t-1}}{x_{t-1}}.
$$

So it is very handy to just difference the log levels to get the growth
rates. Log-difference can also be interpreted as the continuously
compounded rate of change:

$$
\frac{x_t}{x_{t-1}}=e^g \implies g = \ln x_t - \ln x_{t-1}.
$$

Log-difference also has the property of summability: summing up a
sequence of log-differences gives back the log level provided the
initial level. It is not as handy if you want to recover the levels from
percentage growth.

$$
\ln x_t = x_0 + \sum_{j=1}^{t} (\ln x_j - \ln x_{j-1}).
$$
