# ACF and PACF

A time series is notationally represented by
$\{\dots, y_{t-1}, y_t, y_{t+1}, y_{t+2}, \dots\}$ , which is a sequence
of random variables. We think of each variable at a time point $t$ as a
random variable, whose realized value is drawn from some distribution.

A distinguishing feature of this sequence is temporal dependence. That
is, the distribution of $y_t$ conditional on previous value of the
series depends on the outcome of those previous observations. It is of
particular interest how observations are correlated across time. A big
part of the time series analysis is to exploit this correlation.

## Autocorrelation

The temporal dependence is characterized by the correlation between
$y_t$ and its own lags $y_{t-k}$.

::: {#def-acf}
The $k$-th order autocovariance of $y_t$ is defined as

$$\gamma_k = \text{cov}(y_t, y_{t-k}).$$

The $k$-th order autocorrelation is defined as

$$\rho_k = \frac{\text{cov}(y_t, y_{t-k})}{\text{var}(y_t)} = \frac{\gamma_k}{\gamma_0}.$$
:::

If we plot the autocorrelation as a function of the lag length $k$, we
get the autocorrelation function (ACF). Here is an example of the ACF of
China's monthly export growth (log-difference). The lag on the
horizontal axis is counted by seasonal period. Because it is monthly
data, 1 period is 12 months. We can see the autocorrelation is the
strongest for the first two lags. Longer lags are barely significant.
There are spikes with 12-month and 24-month lags, indicating the
seasonality is not fully removed from the series.

```{r}
#| include: false
library(zoo)
```

```{r}
#| echo: true
#| fig-cap: ACF for monthly export growth
#| label: fig-acfexport
data = readRDS("data/md.Rds")
acf(data$Export, main='Autocorrelation')
```

## Partial Autocorrelation

ACF measures the correlation between $y_t$ and $y_{t-k}$ regardless of
their relationships with the intermediate variables
$y_{t-1},y_{t-2},\dots,y_{t-k+1}$. Even if $y_t$ is only correlated with
the first-order lag, it is automatically made correlated with the $k$-th
order lag through intermediate variables. Sometime we are interested in
the correlation between $y_t$ and $y_{t-k}$ partialling out the
influence of intermediate variables.

::: {#def-pacf}
The partial autocorrelation function (PACF) considers the correlation
between the remaining parts in $y_t$ and $y_{t-k}$ after partialling out
the intermediate effect of $y_{t-1},y_{t-2},\dots,y_{t-k+1}$.

$$
\phi_k = \begin{cases}
\text{corr}(y_t, y_{t-1})=\rho_{_1}, \text{ if } k=1;\\
\text{corr}(r_{y_t|y_{t-1},\dots,y_{t-k+1}}, r_{y_{t-k}|y_{t-1},\dots,y_{t-k+1}}), \text{ if } k\geq 2;
\end{cases}
$$

where $r_{y|x}$ means the remainder in $y$ after partialling out the
intermediate effect of $x$.
:::

In practice, $\phi_k$ can be estimated by the regression

$$
y_t = \mu + \phi_1 y_{t-1} + \phi_2 y_{t-2} + \dots + \phi_k y_{t-k} + \epsilon_t.
$$

The estimated coefficient $\hat\phi_k$ is the partial autocorrelation
after controlling the intermediate lags.

```{r}
#| echo: true
#| fig-cap: PACF for monthly export growth
#| label: fig-pacfexport
pacf(data$Export, main='Partial Autocorrelation')
```
