# Estimating Dynamic Multipliers

This section will cover the specifications commonly used to estimate
dynamic causal effect. Just like in cross-sectional analysis, regression
techniques can always be applied as long as the time series are
covariance stationary without an emphasis on causality. However, we pay
special attention to causal inferences, as we are more interested in
understanding the causality rather than mere correlations in most
empirical researches. We start with the case where the structural shock
is directly observed and move on to the cases where the structural
shocks need to be constructed.

## Distributed Lags

The easiest approach to estimate dynamic causal effect is to include
lags in the specification:

$$
y_t = \beta_0\epsilon_t + \beta_1\epsilon_{t-1} + \cdots + \beta_p\epsilon_{t-p} + u_t,
$$

where $\epsilon_t$ is the structural shock, $u_t$ is everything that
otherwise influences $y_t$. Since $\epsilon_t$ happens randomly, we have
$\mathbb E(u_t |\epsilon_{t-j}) = 0$. Thus, the $\beta$s, which capture
the dynamic causal effect, would be consistently estimated by OLS.

Note that we call it a specification, in a sense that the joint
distribution of the random variables is unknown, which distinguishes
itself from the DGP model in @sec-ar-models. But it does not stop us
from uncovering the causal effect, as long as the exogenous condition
holds.

The effect of a unit change in $\epsilon$ on $y$ after $h$ periods,
which is $\beta_{h}$, is also called the $h$-period **dynamic
multiplier**. Sometimes, we are interested in the accumulated effect
over time, $\beta_0+\beta_1+\cdots+\beta_h$, which is called
**cumulative dynamic multiplier**.

Because $u_t$ is the linear combination of all other current and past
shocks, it is likely serially correlated. So HAC standard errors are
required for robust inferences.

::: {#prp-adl}
Assumptions for a consistent estimation of dynamic causal effects with
distributed lag models:

1.  $\epsilon$ is an exogenous shock,
    $\mathbb E(u_t|\epsilon_t,\epsilon_{t-1},...)=0$;
2.  All variables are stationary;
3.  Regular conditions for OLS to work.
:::

To reduce the serial correlations $\{u_t\}$, and also allow for slow
adjustment of $y_t$, we can also include lagged dependent variables in
the specification, which becomes an **autoregressive distributed lag
(ADL)** specification:

$$
y_t = \phi_1 y_{t-1} + \cdots + \phi_p y_{t-p} + \beta_0\epsilon_t + \beta_1\epsilon_{t-1} + \cdots + \beta_p\epsilon_{t-p} + u_t,
$$

or

$$
\phi(L) y_t = \beta(L) \epsilon_t + u_t.
$$

When lags of the dependent variable are included as regressors, strict
exogeneity fails for sure, because
$X=\{y_{t-1},\dots,\epsilon_t, \epsilon_{t-1},\dots\}$ is correlated
with past errors $u_{t-1}$, despite it is uncorrelated with the
contemporary error $u_t$. The OLS is consistent so long as $\{u_t\}$ are
not serially correlated. Otherwise, $u_t$ would be correlated with $X$
through $u_{t-1}$. The serial correlation can be tested with
Durbin-Watson test or Breusch-Godfrey test.

The dynamic causal effect is more convoluted with the ADL specification
though,

$$
\hat\theta(L) = \hat\phi^{-1}(L) \hat\beta(L).
$$

ADL also require truncated lags. $p$ and $q$ are chosen as an increasing
function of the sample size. In general, choosing $p$ and $q$ to be of
order $T^{1/3}$ would be sufficient for consistency.

## Local Projections

Dynamic causal effect can also be estimated by projecting future
outcomes directly on the shock. Jordà (2005) named it **local
projections (LP)**.

$$
y_{t+h} = \theta_h \epsilon_t + u_{t+h}.
$$

By assumption, $\mathbb E(u_{t+h}|\epsilon_t)=0$. So $\hat\theta_h$ is a
consistent estimate of the $h$-period dynamic multiplier. HAC standard
errors are also required in local projections, as $u_{t+h}$ in are
usually serially correlated.

Readers may wonder, since ADL and LP both give consistent estimates of
the dynamic multipliers, what is the difference between them. There are
two obvious differences:

1.  Lagged shocks do not appear in LP specifications as they do in
    distributed lag specifications.
2.  The LP method requires running separate regressions for each $h$.
    The dynamic response $\{\theta_0,\theta_1,\theta_2, \dots\}$ are
    estimated through multiple regressions rather than one.

The error structure is also different. To see this, suppose the DGP is
an MA($\infty$) process

$$
y_t = \epsilon_t + \theta_1\epsilon_{t-1} + \theta_2\epsilon_{t-2} +\cdots
$$

If we estimate it with a DL specification with two lags,

$$
y_t = \beta_0\epsilon_t + \beta_1\epsilon_{t-1} + u_t,
$$

where $u_t = \sum_{j=2}^{\infty}\theta_j\epsilon_{t-j}$. Exogeneity
would ensure $\hat\beta_1 \to \theta_1$.

We can also estimate it with a local projection (suppose we are
interested in the one-step-ahead dynamic multiplier):

$$
y_{t+1} = \psi_1\epsilon_t + u_{t+1}.
$$

Again, we have consistency $\hat\psi_1\to\theta_1$. But the error
structure is different
$u_{t+1} = \epsilon_{t+1} + \sum_{j=2}^{\infty}\theta_j\epsilon_{t-j}$.

Both the DL and LP specifications may include additional control
variables, which can reduce the variance of the residuals and improve
the efficiency of the estimators.

## Example of Observable Exogenous Shocks

Directly observable exogenous shocks are rare. Here we use an example
from Stock and Watson's textbook, which explores the dynamic causal
effect of cold weather on orange juice prices. Cold weather is bad for
orange production. Orange trees cannot withstand freezing temperatures
that last for more than a few hours. Florida accounts for more than 98
percent of U.S. production of frozen concentrated orange juice.
Therefore, the frozen weather in Florida would reduce the supply and
orange juice and raise the price. The dataset includes the number of
freezing degree days in Florida and the average producer price for
orange juice. Cold weather is plausibly exogenous, which allows us the
utilize the regression framework above to estimate the dynamic causal
effect.

```{r}
#| warning: false
#| message: false
#| fig-cap: "Dynamic Effect of Freezing Days on Orange Juice Price"
#| label: fig-frozen
library(AER)
library(dynlm)
library(lmtest)

data("FrozenJuice") # load data

# compute percentage change on price
pchg = 100*diff(log(FrozenJuice[, 'price']))
sample = ts.union(fdd = FrozenJuice[,'fdd'], pchg)

# distributed lag model
mod = dynlm(pchg ~ L(fdd, 0:6), data = sample)
# confidence interval
ci = coefci(mod, vcov. = NeweyWest)

# plot dynamic multiplier
{
  plot(mod$coefficients[-1], # remove intercept
       type = "l", 
       col = 2, 
       ylim = c(-0.4,1), 
       xlab = "Lag", 
       ylab = "Dynamic Multiplier")
  abline(h = 0, lty = 2)
  lines(ci[-1,1], col = 4)
  lines(ci[-1,2], col = 4)
}
```

We can also use local projections. Note that local projections require
estimating multiple regressions. The coefficients from each of the
regressions constitute the dynamic multiplier.

```{r}
#| warning: false
#| message: false
#| fig-cap: "Local Projections of Freezing Days on Orange Juice Price"
#| label: fig-frozenlp
# apply local projection for horizons 0-6
lps = sapply(0:6, function(h) {
  lp = dynlm(L(pchg, -h) ~ fdd, data = sample)
  ci = coefci(lp, vcov. = NeweyWest)
  c(lp$coefficients[-1], ci[-1,]) # remove intercept
}) |> t() # transpose it

# plot the LP coefficients
{
  plot(lps[,'fdd'], 
       type = "l", 
       col = 2, 
       ylim = c(-0.4,1), 
       xlab = "Horizon", 
       ylab = "LP Coefficient")
  abline(h = 0, lty = 2)
  lines(lps[,'2.5 %'], col = 4)
  lines(lps[,'97.5 %'], col = 4)
}
```

## Example of Constructed Structural Shocks

Most structural shocks in economics are not directly observed, such as
monetary policy shocks, or fiscal policy shocks, yet they are of
profound interest of researchers. As we have explained before,
regressing output or inflation on interest rate does not give a
plausible estimation of the causal effect of monetary policy, due to the
endogeneity problem. Thus, we need to isolate the exogenous part of the
monetary policy from observed variables. The method to achieve this is
an active research field in itself. We here demonstrate the monetary
policy shock for China constructed by Das and Song (2023).

The authors utilize the high-frequency price changes of interest rate
swap around the window of monetary policy announcement to approximate
the monetary policy shock. The rationale of this construction is that,
the price of the financial instrument reflects the expected interest
rate by market participants based on the economic conditions. Therefore,
the sudden change of the price in the tiny window of monetary policy
announcement captures the unexpected part of the monetary policy.

```{r}
#| echo: true
#| fig-cap: "Monetary Policy Shock and Lending Rate"
#| label: fig-mpshocks
library(zoo)

mp = readRDS("data/mpshocks.Rds") # monetary policy shock
md = readRDS("data/md.Rds") # monthly data

plot(na.exclude(cbind(mp$shock_1y, md$LoanRate1Y)), 
     main="", xlab="", ylab=c("Loan Rate", "MP Shock"))
```

We estimate the dynamic causal effect of monetary policy shock on
inflation using the constructed MP shocks. It shows that a tightening of
monetary policy implies a gradual cooling down of inflation. The price
level starts to decline roughly half a year after the initial tightening
shock. However, the confidence interval is wide, suggesting an
insignificant estimation of the policy effect. The result does not
provide very strong evidence underlining the effectiveness of monetary
policy to control inflation.

```{r}
#| echo: true
#| fig-cap: "Dynamic response of inflation on monetary policy shocks"
#| label: fig-infmp
sample = na.exclude(cbind(cpi=md$CPI, mp))

# distributed lag model
mod = dynlm(cpi ~ L(shock_1y, 0:12), sample)
# confidence interval
ci = coefci(mod, vcov. = NeweyWest)

# plot dynamic multiplier
{
  plot(mod$coefficients[-1], # remove intercept
       type = "l", 
       col = 2, 
       ylim = c(-1,1.5), 
       xlab = "Lag", 
       ylab = "Dynamic Multiplier")
  abline(h = 0, lty = 2)
  lines(ci[-1,1], col = 4)
  lines(ci[-1,2], col = 4)
}
```
