# Spurious Regression

It is said, all stationary series are alike, but each non-stationary
series is non-stationary in its own way (remember Leo Tolstoy's famous
quote: *all happy families are alike; each unhappy family is unhappy in
its own way.*)

In all previous chapters, we have been working on stationary processes.
We have shown that similar regression techniques and asymptotic results
hold for stationary processes as for $iid$ observations, albeit not
exactly the same. If a time series is not stationary, we transform it to
stationary by taking differences.

This chapter is devoted to study non-stationary time series. Special
attention is given to unit root processes. We will see the theories
involving non-stationary processes are entirely different from those
applied to stationary processes. This makes unit root analysis an rather
independent topic.

The obsession with unit root in academia have faded away in recent
decades (I do not know if this assessment is accurate). Despite the
topic posses immense theoretical interest, it does not seem to provide
proportionate value for applied studies. Nonetheless, the topic is
indispensable for a comprehensive understanding of time series analysis.

We start by pointing out that, it is very dangerous to blindly include
non-stationary variables in a regression. To illustrate this, consider
two random walks:

$$
\begin{aligned}
x_{t} &= x_{t-1} + \epsilon_{t},\quad\epsilon_{t}\overset{iid}\sim N(0,\sigma_X^2)\\
y_{t} &= y_{t-1} + \eta_{t},\quad\eta_{t}\overset{iid}\sim N(0,\sigma_Y^2)
\end{aligned}
$$

$\epsilon_t$ and $\eta_t$ are independent to each other.

```{r}
set.seed(1)

# generate two random walk processes
x = cumsum(rnorm(200))
y = cumsum(rnorm(200))

plot.ts(cbind(x,y), main="Two Random Walks")
```

We would expect the two series completely uncorrelated, as they are two
independent random processes. However, if we regress $y_t$ on $x_t$, we
would likely find a very strong correlation. This is called a **spurious
regression**.

$$
y_t = \alpha + \beta x_t + u_t
$$

```{r}
lm(y ~ x) |> summary()
```

Note that if we difference the two series to stationary, the spurious
correlation disappears.

```{r}
lm(diff(y) ~ diff(x)) |> summary()
```

If we simulate many random walks, we would observe that a large
percentage of the regressions report statistically significant
relationships even though the series are independent.

```{r}
# Simulation parameters
n <- 100         # Number of observations in each random walk
num_sim <- 1000  # Number of simulations
sig_count <- 0   # Counter for significant p-values
pvals <- numeric(num_sim)  # To store p-values

# Loop over simulations
for (i in 1:num_sim) {
  # Generate two independent random walks of length n
  x <- cumsum(rnorm(n))
  y <- cumsum(rnorm(n))
  
  # Run linear regression of y on x
  model <- lm(y ~ x)
  pval <- summary(model)$coefficients[2, 4]
  pvals[i] <- pval
}

# Calculate percentage of simulations with p-value < 0.05
sig_percent <- mean(pvals < 0.05) * 100
print(paste("Percentage of significant regressions:", sig_percent, "%"))

```

This example gives you a quantitative feel for how frequently spurious
results can occur. What's more interesting, however, is that we can
eliminate this spurious strong correlation by including lags of
dependent and independent variables —

```{r}
# Compute lags for x and y
y_lag = dplyr::lag(y)
x_lag = dplyr::lag(x)

# Regression with lags
lm(y ~ y_lag + x + x_lag) |> summary()
```

Non-stationary time series can cause troubles, but they are also
fascinating topics to explore. In what follows, we will focus on two
types of non-stationary processes: trend-stationary processes and unit
root processes, which are the most common types of non-stationary series
we would encounter in economic and finance. Non-stationary series with
exponential growth can be transformed into linear trend, hence is not of
particular interest. We will start with the relatively easy
tread-stationary processes, and spend most of the paragraphs on unit
root processes.
