# Stationarity

## Stationary Process

::: {#def-stationary}
A stochastic process is said to be **strictly stationary** if its
properties are unaffected by a change of time origin. In other words,
the joint distribution at any set of time is not affect by an arbitrary
shift along the time axis.
:::

::: {#def-covstationary}
A stochastic process is called **covariance stationary** (or **weak
stationary**) if its means, variances, and covariances are independent
of time. Formally, a process $\{y_t\}$ is covariance stationary if for
all $t$ it holds that

-   $\mathbb{E}(y_t) = \mu < \infty$;
-   $\text{var}(y_t) = \gamma_{_0} < \infty$;
-   $\text{cov}(y_t,y_{t-k})=\gamma_k$, for $k=1,2,3,\dots$
:::

Stationarity is an important concept in time series analysis. It
basically says the statistical properties of a time series are stable
over time. Otherwise, if the statistical properties vary with time,
statistics estimated from past values, such autocorrelations, would be
much less meaningful. Strict stationarity requires the joint
distribution being stable, that is moments of any order would be stable
over time. In practice, mostly we only care about the first- and
second-order moments, that is means and variances and covariances.
Therefore, covariance stationary is sufficient.

@fig-stationary shows some examples of stationary and non-stationary
time series. Only the first one is stationary (it is generated from
$i.i.d$ normal distribution). The second one is not stationary as its
mean is not constant over time. The third one is not stationary as its
variance is not constant. The last one is not stationary either, because
its covariance is not constant.

```{r}
#| include: false
library(zoo)
```

```{r}
#| echo: false
#| fig-cap: "Stationary and non-stationary time series"
#| label: fig-stationary
x = seq(1,100, 0.1)
par(mfrow=c(2,2), mar=rep(2,4))
plot(rnorm(500), type='l', xlab='', ylab='')
plot(cumsum(.1+rnorm(100)), type='l', xlab='', ylab='')
plot(x, sin(x-50)/(x-50), type='l', xlab='', ylab='')
plot(x, sin(10000/x), type='l', xlab='', ylab='')
```

Real-life time series are rarely stationary. But they can be transformed
to (quasi) stationary by differencing. @fig-stationary-real shows some
examples of the first-order (log) differences of real-life time series.
They more or less exhibit some properties of stationarity, but not
perfectly stationary. The series can be further "stationarized" by
taking a second-order difference. But these examples are acceptable to
be treated as stationary in our models. Even if they are not perfectly
stationary, the model can be thought of being used to "extract" their
stationary properties.

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Stationary and non-stationary time series (real life)"
#| label: fig-stationary-real
md = readRDS("data/md.Rds")
par(mfrow=c(2,2), mar=rep(2,4))
plot(md$CPI, main="CPI", xlab='', ylab='')
plot(md$Export, main="Export", xlab='', ylab='')
plot(md$TBond1Y, main="Interest Rate", xlab='', ylab='')
plot(md$SHSE, main="Stock Market", xlab='', ylab='')
```

## White Noise

White noise is a special stationary process that is an important
building block of many time series models.

::: {#def-whitenoise}
A stochastic process $w_t$ is called **white noise** if its has constant
mean $0$ and variance $\sigma^2$ and no serial correlation
$\text{cov}(w_t, w_{t-k})=0$ for any $k \neq 0$. The white noise process
is denoted as

$$
w_t \sim \text{WN}(0, \sigma^2).
$$
:::

This is the weakest requirement for while noise. It only requires no
serial correlation. We may impose further assumptions. If every $w_t$ is
independent, it becomes independent white noise
$w_t \sim \perp\text{WN}(0, \sigma^2)$. Independence does not imply
identical distribution. If every $w_t$ is independently and identically
distributed, it is called $i.i.d$ white noise,
$w_t \overset{iid}{\sim} \text{WN}(0, \sigma^2)$. If the distribution is
normal, it becomes the most perfect white noise, that is $i.i.d$
Gaussian white noise, $w_t \overset{iid}{\sim} N(0, \sigma^2)$. The
first plot of @fig-stationary is a demonstration of the $i.i.d$ Gaussian
white noise. In most cases, the weakest form of white noise is
sufficient.

::: callout-tip
## Exercise

Prove that a while noise process is stationary.
:::
